Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-11-18 analysis/paper/paper.qmd
split code for tukey results from gis code chunk

```{r}
#| label: spa-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A. Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution). This map include 3 zones. Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). The individual points indicate assemblages. B. Boxplots for  vegitation zones. The individual points indicate artifacts. C. Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). The individual points indicate assemblages. This map include 14 zones. Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). The individual points indicate assemblages. D. Boxplots for  ecoregion zones, excludings zones with no artifacts. The individual points indicate artifacts."
#| fig-width:   5
#| fig-height:  4

@@

eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
summary(eco_aov)
eco_tur<- TukeyHSD(eco_aov)
# plot(eco_tur)

@@

for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
summary(for_aov)
for_tur<- TukeyHSD(for_aov)
# plot(for_tur)

@@

plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
```
We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns. We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship]. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@spa-gis-tcsa-sites), for example. We observed that TCSA values are stonely accociated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`). However, there is no clear functional pattern throughout the zones (@spa-gis-tcsa-sites panel B and D). There is a wide variation in TCSA regardless of the stemmed points density in each zone.

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in forager toolkits. TCSA can be impacted by other factors such as raw materials and portability. -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster). Interestingly, weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box). We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@spa-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.

Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).

\<--!upto here --\>
 We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation. @no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations. Temperatures differ significantly between sites, with a maximum difference of 5℃. Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C). From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site. Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did. In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics here. As we previously noted, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon). THIS DISCUSSION IS INCOMPLETE, WE NEED TO EXPLORE THESE RELATIONSHIPS ABOUT GEOGRAPHICAL VARIATION AND TEMPORAL VARIATION FURTHER, AND IN A NEW WAY THAT IS NOT THE SAME AS THE QSR PAPER, E.G. FOCUS ON THE LGM AS I MENTIONED IN OUR LAST MEETING.

>>>>

```{r}
#| label: fig-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A. Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution). This map include 3 zones. Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). The individual points indicate assemblages. B. Boxplots for  vegitation zones. The individual points indicate artifacts. C. Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). The individual points indicate assemblages. This map include 14 zones. Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). The individual points indicate assemblages. D. Boxplots for  ecoregion zones, excludings zones with no artifacts. The individual points indicate artifacts."
#| fig-width:   5
#| fig-height:  4

@@

@@

plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
```
```{r}
#| label: fig-tukey-eco-region
#| fig-cap: "."
#| fig-width:   5
#| fig-height:  4

#vegetation region#
eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
summary(eco_aov)
eco_tur<- TukeyHSD(eco_aov)
# plot(eco_tur)

#eco region#
for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
summary(for_aov)
for_tur<- TukeyHSD(for_aov)
# plot(for_tur)

@@

We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns. We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship]. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@fig-gis-tcsa-sites), for example. We observed that TCSA values are stonely associated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`). However, there is no clear functional pattern throughout the zones (@fig-gis-tcsa-sites panel B and D). There is a wide variation in TCSA regardless of the stemmed points density in each zone.

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in forager toolkits. TCSA can be impacted by other factors such as raw materials and portability. -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster). Interestingly, weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box). We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. We found no clear correlation between population trend and stemmed points. Population could possibly

Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@fig-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).

Figure 10 from @park2022did

\<--!upto here --\>
 We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation. @no-figure-no-more: B shows that  slowly decreased until around 22 ka and then went up again with a little range of fluctuations. Temperatures differ significantly between sites, with a maximum difference of 5℃. Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C). From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site. Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).

THIS DISCUSSION IS INCOMPLETE, WE NEED TO EXPLORE THESE RELATIONSHIPS ABOUT GEOGRAPHICAL VARIATION AND TEMPORAL VARIATION FURTHER, AND IN A NEW WAY THAT IS NOT THE SAME AS THE QSR PAPER, E.G. FOCUS ON THE LGM AS I MENTIONED IN OUR LAST MEETING.
